---
title: "FC - pi plot"
author: "Meng"
date: "2023-06-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(rattle)
directory="/Users/yuanmeng/Library/CloudStorage/OneDrive-UniversityofToronto/Manuscripts/rumex_haploid_sel/pixy/"
```
## merge with pi data
```{r}
pixy <- read.table(paste0(directory,"pixy_pi_syn.0329.genewise.txt"), header=TRUE) 
pixy <- pixy %>% filter(no_sites>=50) # 19542
colnames(pixy)[2:4] <- c("chrom", "start", "end")
pixy <- inner_join(pixy, gff, by=c("chrom", "start", "end")) # 15049
write.table(pixy, file = paste0(directory,"pixy_filtered.txt"), quote=F) 

DE_tissue<-as.data.frame(result_tissue)
DE_tissue <- DE_tissue %>% mutate(gene = rownames(DE_tissue))
DE_tissue$gene <- str_sub(DE_tissue$gene, end=-4)
DE_tissue <- left_join(DE_tissue,cnt,by="gene")
DE_tissue <- DE_tissue %>% filter(p_mean>5 | l_mean>5)

FC_pixy<- inner_join(DE_tissue,pixy,by=c("gene")) # 11735
write.table(FC_pixy, file = paste0(directory,"FC_pixy_innerjoin.txt"), row.names = FALSE, quote = FALSE)
```


## prepare gene sets
separate to p-biased and l-biased
then 4 quantiles within each category
DE gene cutoffs: padj<0.1 & baseMean >= 1
logFC>0: pollen biased, logFC <-: leaf biased
```{r}
pollen_bias <- FC_pixy %>% filter(log2FoldChange>0) # 4391
leaf_bias <- FC_pixy %>% filter(log2FoldChange<0) # 7344

bin<-binning(pollen_bias$log2FoldChange, bins=4, method="quantile",
                     labels=NULL, ordered=TRUE, weights=NULL)
bin<-binning(leaf_bias$log2FoldChange, bins=4, method="quantile",
                     labels=NULL, ordered=TRUE, weights=NULL)
plot(bin)
```

break and tags for FC
```{r}
breaks <- c(-30,-3.02,-2.16,-1.27,0,1.02,2.54,5.36,30)
# specify interval/bin labels
tags <- c("(,-3.02]","(-3.02,-2.16]", "(-2.16,-1.27]", "(-1.27,0)", "(0,1.02]", "(1.02,2.54]","(2.54,5.36]", "(5.36,)")

# bucketing values into bins
group_tags <- cut(FC_pixy$log2FoldChange, 
                  breaks=breaks, 
                  include.lowest=TRUE, 
                  right=FALSE, 
                  labels=tags)
# inspect bins
summary(group_tags)

ratio_groups <- factor(group_tags, 
                           levels = tags,
                           ordered = TRUE)
```

pollen
1.02]
2.54]
5.36]
26.2]

leaf
-3.02]
-2.16]
-1.27]
-0.101]



## plot
```{r}
v <- FC_pixy %>% select(log2FoldChange,avg_pi) 
vgroup <- as_tibble(v) %>% 
  mutate(tag = case_when(
    log2FoldChange > 0 & log2FoldChange <= 1.02 ~ tags[5],
    log2FoldChange > 1.02 & log2FoldChange <= 2.54 ~ tags[6],
    log2FoldChange > 2.54 & log2FoldChange <= 5.36 ~ tags[7],
    log2FoldChange > 5.36   ~ tags[8],
    
    log2FoldChange <= -3.02 ~ tags[1],
    log2FoldChange > -3.02 & log2FoldChange <= -2.16 ~ tags[2],
    log2FoldChange > -2.16 & log2FoldChange <= -1.27 ~ tags[3],
    log2FoldChange > -1.27 & log2FoldChange < 0 ~ tags[4]

    ))
#summary(vgroup)

vgroup$tag <- factor(vgroup$tag,
                       levels = tags,
                       ordered = FALSE)

ggplot(data = vgroup, mapping = aes(x=tag,y=avg_pi)) + 
    geom_jitter(color="#1F78B4",alpha=0.2) +
    geom_boxplot(fill="#1F78B4",color="black",alpha=0.3) + 
    labs(x='g_s_ratio') +
    guides(color=FALSE) + ylim(0,0.06)+
    theme_classic() + theme(text = element_text(size = 18))

```
 (,-3.02] (-3.02,-2.16] (-2.16,-1.27]     (-1.27,0)      (0,1.02]   (1.02,2.54] 
         1834          1837          1838          1835          1097          1105 
  (2.54,5.36]       (5.36,) 
         1093          1096 

 



